Reproducible, flexible and high throughput data extraction from primary literature: The \pkg{metaDigitise} \proglang{R} package

### 2017-11-20 docs/manuscript.tex
Small changes

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Joel L. Pick, Shinichi Nakagawa, Daniel W.A. Noble\\
  Ecology and Evolution Research Center\\
  \emph{and}\\
  School of Biological, Earth and Environmental Sciences\\
  University of New South Wales\\
  b\\
  b\\
  E-mail: \email{joel.l.pick@gmail.com}\\
  URL: \url{https://eeecon.uibk.ac.at/~zeileis/}
}

@@

\section{Introduction}
In many different contexts, researchers need to make use of data presented in primary literature. Most notably, this includes meta-analysis, which is becoming increasingly common in many research fields. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether particular effects are common across studies and to explain variation among these effects (Nakagawa et al. 2017). Meta-analysis therefore relies foremost on data extracted from primary literature and more specifically statistics that have been reported in the text or tables of research papers. These summary statistics are, however, frequently presented in figures and so need to be manually extracted using digitising programs. Although there are several existing tools to perform tasks like this (EXAMPLES), these tools are not designed this specific purpose (i.e. a meta-analysis). Specifically, they do not differentiate between common plot types that are use to present data, meaning that they require a large amount of downstream data manipulation. For example, many stand alone programs only provide the raw data or values from plots. This means that these data need to be imported into other programs to calculate correlations (if scatter plots) or errors on means (from mean error plots) need to be back calculated by hand to convert them to a common type (e.g., standard errors to standard deviations). Digitising programs also do not easily allow for different grouping of points to be made, and frequently do not integrate metadata (such as variable names, group names and sample sizes) at the time of data extraction. Additionally, existing stand-alone software does not allow easy import into commonly used statistical software (such as \proglang{R}), or try to optimize the research pipelines to facilitate editing and reproducibility. These are major issues because extracting from figures can be an incredibly time-consuming process. Furthermore, although meta-analysis is an important tool in consolidating the data from many studies, relying on transparency of those studies, many of the processes involved in data extraction are opaque and difficult to reproduce, making extending studies problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and go a long way to resolving the reproducibility crises we are seeing in many fields (REF).

Here, we present an interactive \proglang{R} package, \pkg{metaDigitise}, which is designed for large scale data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools specific to data extraction from common plot types (mean and error plots, box plots, scatter plots and histograms, see Figure \ref{fig:plot_type}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. It also provides users with options to conduct the necessary calculations on raw data immediately after extraction so that comparable summary statistics can be obtained quickly. Summaries will condense multiple figures into data frames or lists (depending on the type of figures) and these objects can easily be exported from \proglang{R}, or if using the raw data, analysed in any way the user desires. Conveniently, when needing to process many figures at different times \pkg{metaDigitise} will only import figures not already completed within a directory. This makes it easy to add new figures at any time. \pkg{metaDigitise} has also been built for reproducibility in mind. It has functions that allow users to redraw their digitisations on figures, make corrections and access the raw calibration data which is written automatically for each figure that is digitised into a special folder within the directory. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analysts to update meta-analyses more easily.

>>>>

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Joel L. Pick, Shinichi Nakagawa, Daniel W.A. Noble\\
  Ecology and Evolution Research Centre\\
  \emph{and}\\
  School of Biological, Earth and Environmental Sciences\\
  University of New South Wales\\
  Kensington, NSW 2052\\
  Sydney, AUSTRALIA\\
  E-mail: \email{joel.l.pick@gmail.com}\\
  URL: \url{https://eeecon.uibk.ac.at/~zeileis/}
}

@@

\section{Introduction}
In many different contexts, researchers need to make use of data presented from primary literature. Most notably, this includes meta-analysis, which is becoming increasingly common in many research fields. Meta-analysis uses effect size estimates and their sampling variance, taken from many studies, to understand whether observed effects are common across studies and to explain variation among these effects \citep{Naka2017, Glass1976, Koricheva2013, Borenstein2009}. Meta-analysis therefore relies on statistics that have been reported, for example, in text or tables of research papers. However, summary statistics are very frequently presented in figures, and so, meta-analysts need to manually extract statistical information using digitising programs. Although there are several existing tools to perform these tasks (e.g., graphClick, dataThief), they are not designed for the unique requirements of meta-analysis. Specifically, they do not differentiate between common plot types that can be used to present data, meaning that they require a large amount of downstream data manipulation after digitising, data from figures. For example, many stand alone programs only provide the raw data or specific values from digitised plots. This means that these data need to be imported into other programs to calculate correlations (if scatter plots) or errors on means (from mean error plots). Additionally, errors often need to be converted to usable types (e.g., standard deviations from standard errors) manually. Digitising programs also do not easily distinguish different groups of points within a plot, and frequently do not integrate metadata (such as variable names, group names and sample sizes) during data extraction. Existing stand-alone software also do not allow easy import into commonly used statistical software (such as \proglang{R}), or try to optimize the research pipelines to facilitate editing and reproducibility. These are major issues because extracting from figures can be an incredibly time-consuming process. Furthermore, many of the processes involved in data extraction during a meta-analysis can be opaque and difficult to reproduce, making updating of meta-analyses problematic. Having a tool that facilitates reproducibility in meta-analyses will increase transparency and go a long way to resolving the reproducibility crises we are seeing in many fields \citep{peng_reproducible_2006, peng_reproducible_2011, sandve_ten_2013}.

Here, we present an interactive \proglang{R} package, \pkg{metaDigitise}, which is designed for large scale data extraction from figures, specifically catering to the the needs of meta-analysts. To this end, we provide tools specific to data extraction from common plot types (mean and error plots, box plots, scatter plots and histograms, see Figure \ref{fig:plot_type}). \pkg{metaDigitise} operates within the \proglang{R} environment making data extraction, analysis and export more streamlined. It also provides users with options to conduct the necessary calculations on raw data immediately after extraction so that comparable summary statistics can be obtained quickly. Summaries will condense multiple figures into data frames or lists (depending on the type of figures) and these objects can easily be exported from \proglang{R}, or if using the raw data, analysed in any way the user desires. Conveniently, when needing to process many figures at different times \pkg{metaDigitise} will only import figures not already completed within a directory. This makes it easy to add new figures at any point during the projects lifetime. \pkg{metaDigitise} has also been built for reproducibility in mind. It has functions that allow users to redraw their digitisations on figures, make corrections and access the raw calibration data which is written automatically for each figure that is digitised into a special folder within the directory. This makes sharing figure digitisation and reproducing the work of others simple and easy, and allows meta-analysts to update meta-analyses more easily.
